# This file has the R code use to generate the two figures in the text. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
source("check_packages.R")
source("useful_functions.R")
```

```{r loaddata, include=FALSE}
load("output/final_dataset.RData")


# Code for generating Figure 1

## El Niño and Avian Flu shading data with intensity labels

event_shading <- tribble(
  ~xmin, ~xmax, ~event,
  # Very Strong Events (ONI ≥ 2.0°C)
  1965, 1966, "Very Strong El Niño", # Peak ONI 2.0°C
  1972, 1973, "Very Strong El Niño", # Peak ONI 2.1°C
  1982, 1983, "Very Strong El Niño", # Peak ONI 2.2°C
  1997, 1998, "Very Strong El Niño", # Peak ONI 2.3°C
  2015, 2016, "Very Strong El Niño", # Peak ONI 2.3°C
  2026, 2027, "Very Strong El Niño", # Peak ONI 2.3°C
  
  # Strong Events (ONI ≥ 1.5°C & < 2.0°C)
  1957, 1958, "Strong El Niño",      # Peak ONI 1.8°C
  1987, 1988, "Strong El Niño",      # Peak ONI 1.7°C
  1991, 1992, "Strong El Niño",      # Peak ONI 1.8°C
  2023, 2024, "Strong El Niño",      # Peak ONI 1.7°C
  
  # Other Annotations
  2022 + 10/12, 2023 + 5/12, "Avian Flu"  # Nov 2022 to May 2023
)

# Create scaling factor so birds are visible

max_birds <- max(final_dataset$birds_t, na.rm = TRUE)
max_ssb <- max(final_dataset$ssb, na.rm = TRUE)
scaling_factor <- max_birds / max_ssb

# PLOT
ggplot() +
  
  # Shaded event bands
  geom_rect(
    data = event_shading,
    inherit.aes = FALSE,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf, fill = event),
    alpha = 0.3,
    show.legend = TRUE) +
  
  # Guano birds
  geom_line(data = final_dataset,
            aes(x = year, y = birds_t, color = "Guano Birds"),
            size = 1.1) +
  
  geom_point(data = final_dataset,
             aes(x = year, y = birds_t, color = "Guano Birds"),
             size = 1) +
  
  # Anchovy catch
  geom_line(data = final_dataset,
            aes(x = year, y = anch_catch* scaling_factor, color = "Anchoveta Landings"),
            linetype = 5, size = 0.8) +
#  geom_point(data = final_dataset,
#             aes(x = year, y = anch_catch, color = "Anchoveta Landings"),
#             shape = 24, fill="#6baed6", size = 0.8) +
  
  # SSB BIOMASS
  geom_line(data = final_dataset,
            aes(x = year, y = ssb* scaling_factor, color = "SSB"),
            size = 1.1) +
  geom_point(data = final_dataset,
             aes(x = year, y = ssb* scaling_factor, color = "SSB"),
             size = 1.1) +

# Fix Axes
# x axis
  scale_x_continuous(breaks = c(seq(1910, 2020, 10), 2025), expand = c(0, 0),
  limits = c(1907.5, 2027)) +
# y axes  
scale_y_continuous(name = "Guano Birds (millions)", limits = c(-0.3, 25),
     breaks = seq(0, 25, 2),
     expand = c(0, 0),
sec.axis = sec_axis(
      trans = ~ . / scaling_factor,
      name = "Anchoveta SSB (million t)",
      breaks = seq(0, 18, 2))) +

# Colors
scale_color_manual(name = "Variable",
  breaks = c("Guano Birds", "SSB", "Anchoveta Landings"),
  values = c("Guano Birds" = "#d7191c", "Anchoveta Landings" = "#6baed6", "SSB" = "#2c7bb6")) +
  scale_fill_manual(name = "Event",
    breaks = c("Strong El Niño", "Very Strong El Niño", "Avian Flu"),
    values = c(
      "Strong El Niño" = "#a6dba0",       
      "Very Strong El Niño" = "#008837",
      "Avian Flu" = "#d01c8b")) +
  
  # Theme
  labs(x = "Year", title = "") +
  theme_bw() +
  theme(
    panel.grid = element_blank(),
    legend.title = element_blank(),
    axis.text.x = element_text(angle = 0, hjust = 0.5),
    legend.position = "bottom",
    # Bold axis labels
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold"),
    axis.title.y.right = element_text(face = "bold"),
    # Remove legend key background
    legend.key = element_blank()) +
    guides(color = guide_legend(order = 1, override.aes = list(fill = NA, shape = NA)),
    fill = guide_legend(order = 2, override.aes = list(linetype = 0, shape = NA)))


# Code for generating Figure 2

# Spanish month names
spanish_months <- c("Enero" = 1, "Febrero" = 2, "Marzo" = 3, "Abril" = 4,
                    "Mayo" = 5, "Junio" = 6, "Julio" = 7, "Agosto" = 8,
                    "Septiembre" = 9, "Octubre" = 10, "Noviembre" = 11, 
                    "Diciembre" = 12)

# Prepare data
monthly_data <- aves_guaneras %>%
  rename(Sitio = `Islas y puntas`) %>%
  mutate(
    Mes_num = spanish_months[Mes],
    Fecha = make_date(Año, Mes_num, 1),
    Guanay = `Guanay Reproductivo` + `Guanay No Reproductivo`,
    Piquero = `Piquero Reproductivo` + `Piquero No Reproductivo`,
    Pelicano = `Pelícano Reproductivo` + `Pelícano No Reproductivo`,
    Total_Adults = Guanay + Piquero + Pelicano
  ) %>%
  group_by(Fecha) %>%
  summarise(across(c(Guanay, Piquero, Pelicano, Total_Adults), sum, na.rm = TRUE)) %>%
  filter(Fecha >= "2022-01-01") %>%
  filter(!(year(Fecha) == 2022 & month(Fecha) == 12)) %>%
  pivot_longer(cols = c(Guanay, Piquero, Pelicano), names_to = "Especie", 
               values_to = "Poblacion") %>%
  mutate(Especie = ifelse(Especie == "Pelicano", "Pelican", Especie))

# Event shading
event_shading <- tribble(
  ~xmin, ~xmax, ~event, ~color,
  "2023-06-01", "2024-04-01", "El Niño", "orange",
  "2022-11-01", "2023-05-01", "Avian Flu", "red"
) %>% mutate(across(c(xmin, xmax), as.Date))

# Labels
label_data <- monthly_data %>%
  group_by(Fecha) %>%
  slice(1) %>%
  
  filter(
    (month(Fecha) %% 2 == 0 |
       Fecha %in% as.Date(c("2023-07-01", "2026-03-01", "2022-01-01"))) &
      !Fecha %in% as.Date(c("2023-08-01", "2026-02-01", "2022-04-01", "2022-08-01"))
  ) %>%
  
  mutate(
    angle = 45,
    hjust = -0.1,
    vjust = -1.2,
    
    angle = case_when(
      Fecha == as.Date("2022-01-01") ~ 270,
      Fecha == as.Date("2022-06-01") ~ 0,
      Fecha == as.Date("2022-08-01") ~ 180,
      Fecha == as.Date("2026-03-01") ~ 270,
      Fecha == as.Date("2022-02-01") ~ 270,
      Fecha == as.Date("2022-10-01") ~ 270,
      TRUE ~ angle
    ),
    
    hjust = case_when(
      Fecha == as.Date("2022-01-01") ~ 1.1,
      Fecha == as.Date("2023-02-01") ~ -0.5,
      Fecha == as.Date("2022-06-01") ~ 0.5,
      Fecha == as.Date("2022-08-01") ~ -0.5,
      Fecha == as.Date("2026-03-01") ~ 1.2,
      Fecha == as.Date("2022-02-01") ~ 1.3,
      Fecha == as.Date("2022-10-01") ~ 1.1,
      TRUE ~ hjust
    ),
    
    vjust = case_when(
      Fecha == as.Date("2022-01-01") ~ -0.2,
      Fecha == as.Date("2023-02-01") ~ -2.0,
      Fecha == as.Date("2022-06-01") ~ -1,
      Fecha == as.Date("2022-08-01") ~ 0,
      Fecha == as.Date("2026-03-01") ~ 0.3,
      Fecha == as.Date("2022-10-01") ~ 0.5,
      TRUE ~ vjust
    )
  )

# Plot
ggplot(monthly_data) +
  
  geom_rect(data = event_shading,
            aes(xmin = xmin, xmax = xmax, ymin = 0, ymax = 5500000, fill = event),
            alpha = 0.2) +
  
  geom_col(aes(x = Fecha, y = Poblacion, fill = Especie),
           alpha = 0.8, position = "stack") +
  
  geom_line(aes(x = Fecha, y = Total_Adults),
            color = "black", size = 0.7, linetype = "dashed") +
  
  geom_point(aes(x = Fecha, y = Total_Adults),
             color = "black", size = 1.2) +
  
  geom_text(data = label_data,
            aes(x = Fecha, y = Total_Adults,
                label = scales::comma(Total_Adults),
                angle = angle, hjust = hjust, vjust = vjust),
            size = 2.7, color = "black") +
  
  scale_x_date(
    limits = c(as.Date("2021-12-15"), as.Date("2026-03-15")),
    expand = c(0, 0),
    breaks = seq.Date(
      from = as.Date("2022-01-01"),
      to = as.Date("2026-03-01"),
      by = "1 month"),
    date_labels = "%b %Y") +
  
  scale_y_continuous(
    name = "Adult birds (millions)",
    labels = function(x) x / 1000000,
    limits = c(0, 5500000),
    breaks = seq(0, 5500000, by = 1000000),
    expand = c(0, 0)) +
  
  scale_fill_manual(
    name = "",
    values = c("El Niño" = "orange",
               "Avian Flu" = "red",
               "Guanay" = "#1f77b4",
               "Piquero" = "#ff7f0e",
               "Pelican" = "#2ca02c"),
    breaks = c("Guanay", "Piquero", "Pelican", "Avian Flu", "El Niño")) +
  
  labs(title = "", x = "Date") +
  
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.box = "horizontal",
    legend.key.size = unit(0.5, "cm"),
    legend.spacing.x = unit(0.5, "cm"),
    axis.title.x = element_text(face = "bold", size = 12),
    axis.title.y = element_text(face = "bold", size = 12)) +
  guides(fill = guide_legend(nrow = 1, byrow = TRUE))

